Open-source tools in R for landscape ecology

### 2021-04-26 paper/paper.Rmd
Adding CRS section

---
title: Open-source tools in R for landscape ecology
authors: 
- name: Maximillian H.K. Hesselbarth
  address:
    a) Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
    b) Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
  email: mhk.hesselbarth@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Johannes Signer
  address: Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Buesgenweg 3, 37077 Goettingen, Germany
  email: jsigner@uni-goettingen.de
- name: Laura J. Graham
  address: 
    a) Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
    b) Biodiversity, Ecology & Conservation Group, IIASA, Vienna, Austria 
  email: l.graham@bham.ac.uk
keywords:
- spatial data
- statistical programming language
- "*R* packages"
- reproducibility
- scientific software
abstract: |
  \textbf{Purpose of review} Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years. In particular, the \textit{R} programming language provides a wealth of community developed tools for landscape ecology. \textbf{Recent findings} In this paper, we outline existing packages for downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis. Additionally, we outline the results of a survey of \textit{R} users within the landscape ecology community. \textbf{Summary} We found that landscape ecologists are generally satisfied with the functionality available within \textit{R}, and that as a community they are continually further develop the functionality available. Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools. 
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: |
  \usepackage{caption}
  \usepackage{tabularx}
  \usepackage{multirow}
  \usepackage{placeins}
  \usepackage[modulo]{lineno}
  \usepackage{xcolor}
  \usepackage[none]{hyphenat}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
    citation_package: default
editor_options:
  chunk_output_type: console
  markdown:
    wrap: sentence
---

@@

@@

>>>>

---
title: Open-source tools in R for landscape ecology
authors: 
- name: Maximillian H.K. Hesselbarth
  address:
    a) Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
    b) Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
  email: mhk.hesselbarth@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Johannes Signer
  address: Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Buesgenweg 3, 37077 Goettingen, Germany
  email: jsigner@uni-goettingen.de
- name: Laura J. Graham
  address: 
    a) Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
    b) Biodiversity, Ecology & Conservation Group, IIASA, Vienna, Austria 
  email: l.graham@bham.ac.uk
keywords:
- spatial data
- statistical programming language
- "*R* packages"
- reproducibility
- scientific software
abstract: |
  \textbf{Purpose of review} Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years. In particular, the \textit{R} programming language provides a wealth of community developed tools for landscape ecology. \textbf{Recent findings} In this paper, we outline existing packages for downloading, processing and visualisation of spatial data, as well as those specifically developed for ecological analysis. Additionally, we outline the results of a survey of \textit{R} users within the landscape ecology community. \textbf{Summary} We found that landscape ecologists are generally satisfied with the functionality available within \textit{R}, and that as a community they are continually further develop the functionality available. Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools. 
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: |
  \usepackage{caption}
  \usepackage{tabularx}
  \usepackage{multirow}
  \usepackage{placeins}
  \usepackage{float}
  \usepackage[modulo]{lineno}
  \usepackage{xcolor}
  \usepackage[none]{hyphenat}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
    citation_package: default
editor_options:
  chunk_output_type: console
  markdown:
    wrap: sentence
---

@@

\begin{table}[H]

@@

\textcolor{blue}{Until 2020, spatial \textit{R} objects stored information about their CRS using Proj4 string representations, and these strings could also be used for CRS transformations.
However, the major changes in upstream software, namely PROJ6 and GDAL3, limited the capabilities of the Proj4 strings and started to use a new representation called WKT2-2019 for CRS storing.
Thus, spatial objects saved with degraded Proj4 string representation to \textit{R} file formats could be incorrectly read, and special attention should be given to check their correctness.
Proj4 string representation is also not recommended to set or transform CRS and instead the previously introduced EPSG codes should be used.
A more detailed explanation of the recent CRS changes can be found in \cite{Bivand2020}.}
